Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-16 submission/manuscript.Rmd
add summary paragraph at beg of discussion

# Discussion
We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells and not somatic cells of the testis. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. Female _Kdm5c_-KO EpiLCs expressed over double the number of germline genes than _Kdm5c_-KO males and germline genes dysregulated in both sexes were expressed more highly in females, demonstrating females have increased sensitivity to germline gene dysregulation. KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs. 
- Change: we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

- Thus, systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, will unveil key mechanisms underlying the demarcation between soma and germline identity. These results will provide molecular footholds that can then be exploited to test the contribution of ectopic germline genes on neurodevelopment.

>>>>

# Discussion
In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity. We observed global dysregulation of tissue-enriched genes in the _Kdm5c_-KO brain and EpiLCs, including substantial misexpression of testis, liver, muscle, and ovary-enriched genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells, rather than somatic cells. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. We demonstrated female EpiLCs are more sensitive to germline gene de-repression, as the number germline genes misexpressed and their degree of dysregulation was greater in _Kdm5c_-KO females compared to males. KDM5C is enriched at germline gene promoters in EpiLCs but is bound to only a subset of germline genes transcribed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA methylation at CpG islands for a subset of germline genes in EpiLCs. Therefore, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary. By systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, we unveiled unique regulatory mechansisms governing distinct germline gene classes in somatic lineages. Altogether, these results provide molecular footholds that in the future can be exploited to test the contribution of ectopic germline genes upon neurodevelopment.

---

### Combining sections
